Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-12 Manuscript/manuscript.Rmd
results edits

In this study, we have developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. These include the best combination of predictor type, resampling technique, and data resolution. We utilize random forest classification models built on Hi-C data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three predictor types, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC). We performed 10-fold cross-validation for parameter tuning and to guard against over-fitting. We found that the optimally built cell line specific model is one implemented using distance type predictors, with SMOTE resampling, at the 10-kb data resolution. We also determined that simply including transcription factor binding sites was necessary for optimal performance. The inclusion of chromatin states and histone modifications was found to induce noise into the model. Specifically, we found that the transcription factors CTCF, ZNF143, RAD21, and SMC3 exhibited high predictive importances, confirming there involvement in TAD boundary formation. We then aimed to evaluate the cell line specificity of TAD boundaries by validating our optimally built models across cell lines. We found that models built and validated on the same cell line outperformed models that were validated across cell lines, confirming the cell line specific nature and plasticity of TAD boundaries and functional genomic annotations respectively. <!--[add conclusion from results of predicting finer bins]-->

@@

## USING OPTIMALLY BUILD CLASSIFIER TO PREDICT BOUNDARIES AT FINER RESOLUTION

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First, we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

We sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were found to be specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution [??? spell out resolutions exactly.] respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically. This indicates a necessity for computationally specific models [??? Do not understand] in predicting TAD boundary formations.

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.  [??? It is low because the upper bound is 0.5. It is an unconventional measure.]

@@

A total of 107 and 152 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line, respectively. [??? Is that so? How many are common?] To establish the feature space used for classification, we constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

>>>>

In this study, we have developed a computational framework aimed at determining the optimal set of data level characteristics necessary for optimal prediction of TAD boundaries and identification of influential genomic annotations. These include the best combination of predictor type, resampling technique, and data resolution. We utilize random forest classification models built on Hi-C data from two cell lines, including GM12878 and K562, obtained at four different resolutions, including 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For each cell line and resolution, we compared three predictor types, including counts, percents, and distances. Additionally, we compared three re-sampling techniques, including random under-sampling, random over-sampling, and a weighted combination of both referred to as Synthetic Minority Over-sampling Technique (SMOTE). Model performances were evaluated and compared using Matthews Correlation Coefficient (MCC). We performed 10-fold cross-validation for parameter tuning and to guard against over-fitting. We found that the optimally built cell line specific model is one implemented using distance type predictors, with SMOTE resampling, at the 10-kb data resolution. We also determined that simply including transcription factor binding sites was necessary for optimal performance. The inclusion of chromatin states and histone modifications was found to induce noise into the model. Specifically, we found that the transcription factors CTCF, ZNF143, RAD21, and SMC3 exhibited high predictive importances, confirming there involvement in TAD boundary formation. We then aimed to evaluate the cell line specificity of TAD boundaries by validating our optimally built models across cell lines. We found that models built and validated on the same cell line outperformed models that were validated across cell lines, confirming the cell line specific nature and plasticity of TAD boundaries and functional genomic annotations respectively. <!--[add conclusion from results of predicting finer bins]-->

@@

## USING OPTIMALLY BUILT CLASSIFIER TO PREDICT BOUNDARIES AT FINER RESOLUTION

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First, we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

We sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were found to be specific to one of the two cell lines (Figure 2B). It was found that X%, 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, while X%, 27%, 27%, 29%, and 33% of TAD boundary regions were specific to the K562 cell line across each of the 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively. Given the plasticity in boundary location, we opted to build cell line specific models for predicting TAD boundary regions.

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as twice the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.

@@

A total of 78 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models. We constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) were defined as the total number of peak regions that overlapped within a given genomic bin (Figure 3A). Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by a peak region (Figure 3B). **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations (Figure 3C). A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). Each predictor type provides a granular way to draw associations between functional genomic annotations and the spatial structure of the genome.
